The Galactic Midplane is Not a Plane: Implications for dynamical analysis with {\em Gaia} data and beyond

### 2019-04-22 paper/ms.tex
Cleaning up from mm's changes

\section{Introduction} \label{sec:intro}
Our understanding of the Milky Way is currently undergoing a
revolution 
%mm in the era of {\em Gaia} DR2. 
     as a result of {\em Gaia} Data Release 2 (DR2).  Recent major discoveries include the remnants of a
major merger \citep{2018ApJ...860L..11K, 2018Natur.563...85H,
2018arXiv180704290L, 2019MNRAS.482.3426M}, a phase-space ``spiral'' in the
solar neighborhood \citep{2018Natur.561..360A} possibly indicating either
local substructure infall \citep{2018MNRAS.481.1501B, 2018arXiv180800451L} or
bar buckling \citep{2018arXiv181109205K}, and a gap suggestive of perturbation
by a dark matter subtructure in the tidal stream GD1
\citep{2018ApJ...863L..20P, 2018arXiv181103631B}. These discoveries all
indicate that the Milky Way's stellar distribution, which demonstrably departs
significantly from axisymmetry, is undergoing phase mixing and dynamical
interactions across a range of spatial and temporal scales.

Underlying the quantitative analysis of many of the mechanisms at work that
give rise to these signatures is the assumption of a global, axisymmetric
Galactocentric coordinate system \citep{2008gady.book.....B} in which the
Sun's position (and therefore the origin of the coordinate system) can be
defined and measured both precisely and accurately. In this system, the center
of the Galaxy is usually taken to be the location of the central supermassive
black hole, Sgr~A\textsuperscript{*} \citep[e.g.][]{2004ApJ...616..872R}. From
stellar motions near Sgr~A\textsuperscript{*}, the distance from the Sun to
the Galactic center
%mm is taken to be $\sim 8.1\text{--}8.3\,\kpc$
    can be measured
\citep{2009ApJ...692.1075G,2018AA...615L..15G}, with a recent 
%mm $\sim0.3\%$ measurement from the GRAVITY collaboration
    measurement using near-infrared interferometry of $8.178 \pm
    0.035$~kpc 
\citep{2019arXiv190405721A}. The Sun's vertical distance
from the Galactic midplane, the other independent measurement that sets this
coordinate system, can be determined by identifying where the stellar density
and velocities reach a maximum (effectively the median height of all disk
stars), and is usually taken to be $\sim 25\,\pc$ \citep{2001ApJ...553..184C},
with more recent measurements from {\em Gaia} DR2 placing it at $\sim 20\,\pc$
\citep{2019MNRAS.482.1417B}. Another strategy is to use the cold gas or HII
regions in the disk to define the Galactic midplane, leading to slightly
different values ($\sim 5\,\pc$) for the Sun's relative height
\citep[e.g.][]{2019ApJ...871..145A}. A pre-{\em Gaia} review of these
measurements is given by \citet{2016ARAA..54..529B}.

Once a Galactocentric coordinate system has been established and a
six-dimensional (6D) phase
space measurement of a star has been made, it is often desirable to convert
this measurement into action space to concisely summarize its projected orbit,
model the stellar distribution function, or find stars with similar dynamical
properties. Actions are conserved quantities that describe the orbit of a
star under the assumption of regular, bound orbits in a system where the
equations of motion are separable in a particular coordinate system. They are
the 
%mm typical
cyclical integral of the canonical momentum  over its conjugate position,
%mm and are given by,
\beq\label{eq:actions}
J_i \equiv
\frac{1}{2\pi} \oint_{\text{orbit}}p_i\,\text{d}x_i\text{,.}
\eeq
Under the assumption of axisymmetry, $i=R,z,\phi$ are the radial, vertical, and azimuthal coordinates
respectively in a cylindrical coordinate system and the $p_i$ are the
conjugate momenta.\footnote{We 
%mm typically [there are no other J_R's in manuscript]
write $J_r$ instead of $J_R$ for
aesthetic reasons.} In a slowly-evolving axisymmetric potential, these actions
are invariant and $J_{\phi} \equiv L_z$
\citep{2008gady.book.....B,2014RvMP...86....1S}.

@@

The transformation to action space is nonlinear in both the phase-space
position and the potential, and requires relatively strong assumptions about
the system's symmetry. 
%mm For instance, 
    However, 
the Galactic potential is not strictly
axisymmetric; with the vast improvement in the quality of phase-space
measurements due to {\em Gaia}, this assumption appears increasingly
inadequate \citep[e.g.][]{2018Natur.561..360A}. Even if this
assumption 
%mm is [counterfactual subjunctive, I believe]
     were
close enough for many purposes, the parameters used in axisymmetric models may
be inadequately constrained by current observations.

In this work we discuss how the imposition of a global axisymmetric coordinate
system based on a local measurement of the midplane location and
Galactocentric distance introduces systematic error in the computation of
actions, especially for stars in the thin disk. Our main point is that the
``local'' Galactic midplane is not the same as the ``global'' Galactic
midplane. For stars relatively far from the Sun, the local midplane location
will not agree with our local midplane extrapolated onto their position,
thanks to relatively small-scale variations in the density of the gas and
stellar distribution. Converting positions and velocities of more distant
stars from a heliocentric to a Galactocentric coordinate system introduces a
systematic bias in the $z$ coordinate, and hence the actions computed for
these stars. The further from the solar neighborhood the target star, the more
likely the mismatch will result in large systematic uncertainty, especially in
the vertical action $J_z$. Depending on the type of orbit being studied, this
sets a limit on the ``dynamical solar neighborhood'' over which the local
coordinate frame can be usefully extended for action-space analysis. A similar
argument applies to 
%mm [given that this is down to 0.3% error, seems a bit academic]
   any remaining 
uncertainty in measurements of 
%mm $R_{\odot}$\footnote{Here,
%and throughout, we use $R_{\odot}$ to mean the distance from the Sun to the
%Galactic center.}
     the Galactocentric radius of the Sun $R_{\odot}$
and to variations in the local standard of rest (LSR).

In Section~\ref{sec:ref_frame}, we describe the general impact coordinate
system errors have on the measured actions. Lacking an empirical, nonlocal
definition of the global Galactic midplane or a measurement of the expected
midplane variation in the Milky Way itself,  in Section~\ref{sec:local_fire}
we 
%mm attempt to quantify the 
    examine the expected 
azimuthal dependence of the local midplane, and
hence the expected size of the dynamical solar neighborhood, using
cosmological, hydrodynamical, zoom-in simulations of Milky Way-mass galaxies
from the Feedback in Realistic Environments (FIRE)
collaboration\footnote{\url{https://fire.northwestern.edu}}
\citep{2016ApJ...827L..23W,2017arXiv170206148H}.

@@

In Figure~\ref{fig:Jz_hist} we plot a histogram of the observed values of
$J_z$ for the thick-disk orbit ({\em upper panel}) and thin-disk orbit ({\em
lower panel}) assuming a $z$ offset of $100\,\pc$ (i.e. the {\em right} panel
of Figure~\ref{fig:one_orbit_wrong_ref}). The true value of $J_z$ is plotted
as a vertical dashed line. Here we see that the 
%mm spread 
    errors 
in $J_z$ induced by a
systematic offset in $z$ are non-Gaussian and bimodal. Furthermore, neither of
the modes lie %mm at the true value of $J_z$.
   at zero. In other words, neither of the modes of the distribution of observed $J_z$ values are centered on the true $J_z$ (i.e. the $J_z$ computed with no offset in $z$).
In the case of the thin-disk orbit
({\em lower panel}), we see that, in addition to the prior complications, the
error distribution is not even centered on 
%mm the true value. 
    null. 
This comes about
when the midplane error is $\gtrsim z_{\text{max}}$, and can be understood by
realizing that in such a case the midplane offset always acts to increase the
value of $J_z$, resulting in a biased estimate of $J_z$.

@@

\item We pointed out that the error distribution of the actions induced by a
coordinate system error is very non-Gaussian. The distribution is bimodal with
{\em neither of the modes peaking at 
%mm the true value}.
     null}. 
As a result, error
propagation is 
%mm very involved 
     complex
when considering actions.

@@

\acknowledgments
We would like to thank Megan Bedell, Tobias Buck, and Todd Phillips
%mm [moving to author list] David W. Hogg, and Kathryn
%Johnston 
%for many useful discussions. A.B. would like to thank Todd Phillips
for helpful discussions. A.B. was supported in part by the Roy \& Diana
Vagelos Program in the Molecular Life Sciences and the Roy \& Diana Vagelos
Challenge Award. 
%mm The work of \Gus{everyone} is supported by the Simons
%Foundation. \Gus{remove last sentence if everyone CCA affiliated}
M.-M.M.L. was partly supported by NSF grant AST18-15461.

>>>>

\section{Introduction} \label{sec:intro}
Our understanding of the Milky Way is currently undergoing a
revolution 
     as a result of {\em Gaia} Data Release 2 (DR2).  Recent major discoveries include the remnants of a
major merger \citep{2018ApJ...860L..11K, 2018Natur.563...85H,
2018arXiv180704290L, 2019MNRAS.482.3426M}, a phase-space ``spiral'' in the
solar neighborhood \citep{2018Natur.561..360A} possibly indicating either
local substructure infall \citep{2018MNRAS.481.1501B, 2018arXiv180800451L} or
bar buckling \citep{2018arXiv181109205K}, and a gap suggestive of perturbation
by a dark matter subtructure in the tidal stream GD1
\citep{2018ApJ...863L..20P, 2018arXiv181103631B}. These discoveries all
indicate that the Milky Way's stellar distribution, which demonstrably departs
significantly from axisymmetry, is undergoing phase mixing and dynamical
interactions across a range of spatial and temporal scales.

Underlying the quantitative analysis of many of the mechanisms at work that
give rise to these signatures is the assumption of a global, axisymmetric
Galactocentric coordinate system \citep{2008gady.book.....B} in which the
Sun's position (and therefore the origin of the coordinate system) can be
defined and measured both precisely and accurately. In this system, the center
of the Galaxy is usually taken to be the location of the central supermassive
black hole, Sgr~A\textsuperscript{*} \citep[e.g.][]{2004ApJ...616..872R}. From
stellar motions near Sgr~A\textsuperscript{*}, the distance from the Sun to
the Galactic center
    can be measured
\citep{2009ApJ...692.1075G,2018AA...615L..15G}, with a recent 
    measurement using near-infrared interferometry of $8.178 \pm
    0.035$~kpc 
\citep{2019arXiv190405721A}. The Sun's vertical distance
from the Galactic midplane, the other independent measurement that sets this
coordinate system, can be determined by identifying where the stellar density
and velocities reach a maximum (effectively the median height of all disk
stars), and is usually taken to be $\sim 25\,\pc$ \citep{2001ApJ...553..184C},
with more recent measurements from {\em Gaia} DR2 placing it at $\sim 20\,\pc$
\citep{2019MNRAS.482.1417B}. Another strategy is to use the cold gas or HII
regions in the disk to define the Galactic midplane, leading to slightly
different values ($\sim 5\,\pc$) for the Sun's relative height
\citep[e.g.][]{2019ApJ...871..145A}. A pre-{\em Gaia} review of these
measurements is given by \citet{2016ARAA..54..529B}.

Once a Galactocentric coordinate system has been established and a
six-dimensional (6D) phase
space measurement of a star has been made, it is often desirable to convert
this measurement into action space to concisely summarize its projected orbit,
model the stellar distribution function, or find stars with similar dynamical
properties. Actions are conserved quantities that describe the orbit of a
star under the assumption of regular, bound orbits in a system where the
equations of motion are separable in a particular coordinate system. They are
the 
cyclical integral of the canonical momentum  over its conjugate position,
\beq\label{eq:actions}
J_i \equiv
\frac{1}{2\pi} \oint_{\text{orbit}}p_i\,\text{d}x_i\text{,.}
\eeq
Under the assumption of axisymmetry, $i=R,z,\phi$ are the radial, vertical, and azimuthal coordinates
respectively in a cylindrical coordinate system and the $p_i$ are the
conjugate momenta.\footnote{We 
write $J_r$ instead of $J_R$ for
aesthetic reasons.} In a slowly-evolving axisymmetric potential, these actions
are invariant and $J_{\phi} \equiv L_z$
\citep{2008gady.book.....B,2014RvMP...86....1S}.

@@

The transformation to action space is nonlinear in both the phase-space
position and the potential, and requires relatively strong assumptions about
the system's symmetry. 
    However, 
the Galactic potential is not strictly
axisymmetric; with the vast improvement in the quality of phase-space
measurements due to {\em Gaia}, this assumption appears increasingly
inadequate \citep[e.g.][]{2018Natur.561..360A}. Even if this
assumption 
     were
close enough for many purposes, the parameters used in axisymmetric models may
be inadequately constrained by current observations.

In this work we discuss how the imposition of a global axisymmetric coordinate
system based on a local measurement of the midplane location and
Galactocentric distance introduces systematic error in the computation of
actions, especially for stars in the thin disk. Our main point is that the
``local'' Galactic midplane is not the same as the ``global'' Galactic
midplane. For stars relatively far from the Sun, the local midplane location
will not agree with our local midplane extrapolated onto their position,
thanks to relatively small-scale variations in the density of the gas and
stellar distribution. Converting positions and velocities of more distant
stars from a heliocentric to a Galactocentric coordinate system introduces a
systematic bias in the $z$ coordinate, and hence the actions computed for
these stars. The further from the solar neighborhood the target star, the more
likely the mismatch will result in large systematic uncertainty, especially in
the vertical action $J_z$. Depending on the type of orbit being studied, this
sets a limit on the ``dynamical solar neighborhood'' over which the local
coordinate frame can be usefully extended for action-space analysis. A similar
argument applies to 
%mm [given that this is down to 0.3% error, seems a bit academic]
   any remaining 
uncertainty in measurements of 
     the Galactocentric radius of the Sun $R_{\odot}$
and to variations in the local standard of rest (LSR).

In Section~\ref{sec:ref_frame}, we describe the general impact coordinate
system errors have on the measured actions. Lacking an empirical, nonlocal
definition of the global Galactic midplane or a measurement of the expected
midplane variation in the Milky Way itself,  in Section~\ref{sec:local_fire}
we 
    examine the expected 
azimuthal dependence of the local midplane, and
hence the expected size of the dynamical solar neighborhood, using
cosmological, hydrodynamical, zoom-in simulations of Milky Way-mass galaxies
from the Feedback in Realistic Environments (FIRE)
collaboration\footnote{\url{https://fire.northwestern.edu}}
\citep{2016ApJ...827L..23W,2017arXiv170206148H}.

@@

In Figure~\ref{fig:Jz_hist} we plot a histogram of the observed values of
$J_z$ for the thick-disk orbit ({\em upper panel}) and thin-disk orbit ({\em
lower panel}) assuming a $z$ offset of $100\,\pc$ (i.e. the {\em right} panel
of Figure~\ref{fig:one_orbit_wrong_ref}). The true value of $J_z$ is plotted
as a vertical dashed line. Here we see that the 
    errors 
in $J_z$ induced by a
systematic offset in $z$ are non-Gaussian and bimodal. Furthermore, neither of
the modes lie
   at zero. In other words, neither of the modes of the distribution of observed $J_z$ values are centered on the true $J_z$ (i.e. the $J_z$ computed with no offset in $z$).
In the case of the thin-disk orbit
({\em lower panel}), we see that, in addition to the prior complications, the
error distribution is not even centered on 
    null. 
This comes about
when the midplane error is $\gtrsim z_{\text{max}}$, and can be understood by
realizing that in such a case the midplane offset always acts to increase the
value of $J_z$, resulting in a biased estimate of $J_z$.

@@

\item We pointed out that the error distribution of the actions induced by a
coordinate system error is very non-Gaussian. The distribution is bimodal with
{\em neither of the modes peaking at 
     null}. 
As a result, error
propagation is 
     complex
when considering actions.

@@

\acknowledgments
We would like to thank Megan Bedell, Tobias Buck, and Todd Phillips
A.B. would like to thank Todd Phillips
for helpful discussions. A.B. was supported in part by the Roy \& Diana
Vagelos Program in the Molecular Life Sciences and the Roy \& Diana Vagelos
Challenge Award. 
%mm The work of \Gus{everyone} is supported by the Simons
%Foundation. \Gus{remove last sentence if everyone CCA affiliated}
M.-M.M.L. was partly supported by NSF grant AST18-15461.
